---
title: "Logical Connectives"
author: "Mika Braginsky"
date: "June 13, 2017"
output: 
  html_notebook: 
    highlight: tango
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = TRUE,
                      dev = "cairo_pdf")
library(tidyverse)
library(magrittr)
library(purrrlyr)
library(glue)
library(forcats)
library(langcog)
theme_set(theme_mikabr())
```

```{r ops}
# meaning = one row of meanings df
meanings <- data_frame(
  "A" = c(FALSE, FALSE, TRUE, TRUE),
  "B" = c(FALSE, TRUE, FALSE, TRUE)
)

# lexicon = subset of columns of operators df
operators <- meanings %>%
  transmute(
    "NO" = FALSE,
    "AND" = A & B,
    "NOT_IMPLIES" = A & !B,
    "A" = A,
    "NOT_CONV_IMPLIES" = !A & B,
    "B" = B,
    "XOR" = xor(A, B),
    "OR" = A | B,
    "NOR" = !(A | B),
    "XNOR" = !xor(A, B),
    "NOT_B" = !B,
    "IMPLIES" = A | !B,
    "NOT_A" = !A,
    "CONV_IMPLIES" = !A | B,
    "NAND" = !(A & B),
    "YES" = TRUE
  )

namify_val <- function(names, val) {
  val %>% rep(length(names)) %>% set_names(names)
}

default_params <- list(
  meaning_prior = function(meaning) 1 / nrow(meanings),
  utterance_cost = namify_val(names(operators), 0),
  utterance_weight = namify_val(names(operators), 1),
  lambda = 3
)
```

A literal listener $L_0$ interprets an utterance $u$ using their prior over meanings $m$ and the entry for the word in their lexicon $\mathcal{L}$:

$$ L_0(m|u,\mathcal{L}) \propto \mathcal{L}(m,u) \cdot P(m)$$

```{r}
listener_literal <- function(lexicon, params) {
  function(meaning, utterance) {
    
    # set of meanings consistent with utterance
    utterance_meanings <- meanings[which(lexicon[[utterance]]),]
    
    p_meaning <- function(m_i) {
      valid <- utterance_meanings %>% filter(A == m_i$A, B == m_i$B) %>% nrow()
      if (valid) params$meaning_prior(m_i) else 0
    }
    
    total <- meanings %>% by_row(p_meaning, .collate = "cols") %$%
      .out %>% sum(na.rm = TRUE)
    log(p_meaning(meaning)) - log(total)
  }
}
```

A pragmatic speaker chooses a utterance $u$ given an meaning $m$ to which they want to refer such that they soft-maximize the probability that the listener will assign to that meaning:

$$ S_n(u|m,\mathcal{L}) \propto \text{exp} (\lambda (\log{L_{n-1}(m|u,\mathcal{L}) - \text{cost}(u)})) $$

```{r}
# level n pragmatic speaker's log posterior of utterance given meaning and lexicon
# n ∈ {1, 3, 5, ...}
speaker_n <- function(n, lexicon, params) {
  function(meaning, utterance) {
    p_utterance <- function(u_i) {
      params$lambda * (listener_n(n - 1, lexicon, params)(meaning, u_i) -
                         params$utterance_cost[u_i])
    }
    total <- sum(exp(map_dbl(names(lexicon), p_utterance)), na.rm = TRUE)
    if (total == 0) -Inf else p_utterance(utterance) - log(total)
  }
}
```

A pragmatic listener, in turn, uses Bayes' rule to in invert the speaker's decision rule:

$$ L_n(m|u,\mathcal{L}) \propto S_{n-1}(u|m,\mathcal{L}) \cdot P(m) $$

```{r}
# level n pragmatic listener's log posterior of meaning given utterance and lexicon
# n ∈ {0, 2, 4, ...}
listener_n <- function(n, lexicon, params) {
  if (n == 0) {
    listener_literal(lexicon, params)
  } else {
    function(meaning, utterance) {
      p_meaning <- function(m_i) {
        speaker_n(n - 1, lexicon, params)(m_i, utterance) + 
          params$meaning_prior(m_i)
      }
      total <- meanings %>% by_row(p_meaning, .collate = "cols") %$%
        .out %>% exp() %>% sum(na.rm = TRUE)
      p_meaning(meaning) - log(total)
    }
  }
}
```

```{r}
type_interp <- function(type, lexicon, params, utterance) {
  if (type == "literal") {
    interp_fun <- listener_literal(lexicon, params)
  } else if (type == "pragmatic") {
    interp_fun <- listener_n(2, lexicon, params)
  }
  meanings %>%
    by_row(~interp_fun(.x, utterance), .collate = "cols") %>%
    rename(p = .out) %>%
    mutate(p = exp(p),
           utterance = utterance,
           meaning = sprintf("A = %d\nB = %d", A, B),
           type = type)
}

interp <- function(lexicon, params) {
  map_df(c("literal", "pragmatic"),
                 function(t) map_df(names(lexicon),
                                    ~type_interp(t, lexicon, params, .x)))
}

plot_interp <- function(interp) {
  ggplot(typical_interp, aes(x = meaning, y = p)) +
    facet_grid(type ~ utterance) +
    geom_bar(stat = "identity") +
    labs(x = "meaning", y = "P(meaning | utterance)",
         title = "Listener interpretation")
}

typical_lexicon <- operators %>% select(AND, OR)
typical_interp <- interp(typical_lexicon, default_params)
```

```{r}
ggplot(typical_interp, aes(x = meaning, y = p)) +
  facet_grid(type ~ utterance) +
  geom_bar(stat = "identity") +
  labs(x = "meaning", y = "P(meaning | utterance)",
       title = "Listener interpretation")
```

```{r}
get_lexicon_eval <- function(lexicon, params) {
  names(lexicon) %>%
    map_df(function(utt) {
      meanings %>%
        by_row(~speaker_n(1, lexicon, params)(.x, utt), .collate = "cols") %>%
        rename(speaker_p = .out) %>%
        by_row(~listener_n(2, lexicon, params)(.x, utt), .collate = "cols") %>%
        rename(listener_p = .out) %>%
        mutate(utterance = utt)
    })
}

lexicon_score <- function(lexicon_eval, params) {
  lexicon_eval %>%
    mutate(prior = params$meaning_prior(list(A = A, B = B)),
           weight = params$utterance_weight[utterance]) %$%
    sum(exp(speaker_p + listener_p) * prior * weight, na.rm = TRUE)
}
```

```{r}
# evaluate all operator inventories of size k
get_inventories_eval <- function(k, params) {
  inventories <- combn(names(operators), k, simplify = FALSE)
  # inv <- data_frame(inventory = inventories) %>%
  #   mutate(lexicon_eval = get_lexicon_eval(operators %>% select_(.dots = inventory),
  #                                          params))
  1:length(inventories) %>%
    map_df(function(i) {
      print(glue("inventory {i} out of {length(inventories)}"))
      get_lexicon_eval(operators %>% select_(.dots = inventories[[i]]), params) %>%
        mutate(i = i) #%>%
    })
    # group_by(i) %>%
    # nest() %>%
    # mutate(score = map_dbl(data, ~lexicon_score(.x, params))) %>%
    # unnest()
}

get_inventories_scores <- function(inventories_eval, params) {
  inventories_eval %>%
    
    group_by(i) %>%
    nest() %>%
    mutate(score = map_dbl(data, ~lexicon_score(.x, params))) %>%
    unnest() %>%
    
    distinct(i, utterance, score) %>%
    group_by(i) %>%
    mutate(op = sprintf("operator_%s", 1:n())) %>%
    spread(op, utterance) %>%
    ungroup() %>%
    mutate_if(is.character, as_factor)
}
```

```{r}
inventories_2_eval <- get_inventories_eval(2, default_params)
inventories_2_scores <- get_inventories_scores(inventories_2_eval)

inventories_2_scores %>%
  select(operator_1, operator_2, score) %>%
  arrange(desc(score))
```

```{r, fig.width=4, fig.height=4}
ggplot(inventories_2_scores, aes(x = operator_1, y = operator_2)) +
  coord_equal() +
  geom_tile(aes(fill = score)) +
  scale_fill_gradient(low = solarized_palette(2)[1],
                      high = solarized_palette(2)[2]) +
  labs(x = "", y = "") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
inventories_3_eval <- get_inventories_eval(3, default_params)
inventories_3_scores <- get_inventories_scores(inventories_3_eval, default_params)

inventories_3_scores %>%
  select(operator_1, operator_2, operator_3, score) %>%
  arrange(desc(score))
```

```{r}
# non-uniform prior over meanings such that predicates tend to be false
# p = parameter of how likely any predicate is to be true
meaning_prior_params <- function(p) {
  params <- default_params
  params$meaning_prior <- function(meaning) {
    num_true <- meaning$A + meaning$B
    p ^ num_true * (1 - p) ^ (2 - num_true)
  }
  params
}
```

```{r}
skewed_3_eval <- get_inventories_eval(3, meaning_prior_params(0.1))
skewed_3_scores <- get_inventories_scores(skewed_3_eval)

skewed_3_scores %>%
  select(operator_1, operator_2, operator_3, score) %>%
  arrange(desc(score))
```

```{r}
# ops <- list(
#   "NO" = function(A, B) FALSE,
#   "AND" = function(A, B) A & B,
#   "NOT_IMPLIES" = function(A, B) A & !B,
#   "A" = function(A, B) A,
#   "NOT_CONV_IMPLIES" = function(A, B) !A & B,
#   "B" = function(A, B) B,
#   "XOR" = function(A, B) xor(A, B),
#   "OR" = function(A, B) A | B,
#   "NOR" = function(A, B) !(A | B),
#   "XNOR" = function(A, B) !xor(A, B),
#   "NOT_B" = function(A, B) !B,
#   "IMPLIES" = function(A, B) A | !B,
#   "NOT_A" = function(A, B) !A,
#   "CONV_IMPLIES" = function(A, B) !A | B,
#   "NAND" = function(A, B) !(A & B),
#   "YES" = function(A, B) TRUE
#   )
# 
# # reduces <- function(operator, expression) {
# #   meanings %>% mutate(op = ops[[operator]](A, B))
# # }
# 
# # given two vectors of inputs, return operator's output
# operate <- function(operator, A, B) {
#   map2_lgl(A, B, ~ops[[operator]](.x, .y))
# }
# 
# function(primitives, inputs = c("A", "B")) {
#   
# }
# 
# operator_weight <- function(operator, primitives) {
#   return(1)
# }
```

```{r}
library(rPython)
python.load("Gates.py")

operator_k <- function(primitives, inputs = c("A", "B")) {
  python.assign("primitives", primitives)
  python.assign("inputs", inputs)
  results <- python.call("min_circuits", primitives, inputs)
  gates <- python.get("gates")
  gate_names <- gates[names(results)]
  results %>% set_names(gate_names) %>% transpose() %>% .[[1]] %>% unlist()
}

operator_k(c("AND", "OR", "NOT_A", "NOT_B"))
#operator_k(c("AND"))
```

```{r}
utterance_weight_params <- function(primitives, inputs = c("A", "B")) {
  params <- default_params
  k <- operator_k(primitives, inputs)
  k_norm <- 1 - k / sum(k)
  params$utterance_weight <- k_norm
  params
}
```

```{r}
MDL_params <- utterance_weight_params(primitives = c("AND", "OR", "NOT_A", "NOT_B"),
                                      inputs = c("A", "B", "YES", "NO"))
#MDL_3_eval <- get_inventories_eval(3, MDL_params)
MDL_3_scores <- get_inventories_scores(inventories_3_eval, MDL_params)

MDL_3_scores %>%
  select(operator_1, operator_2, operator_3, MDL_score = score) %>%
  left_join(inventories_3_scores %>%
              select(operator_1, operator_2, operator_3, score)) %>%
  arrange(desc(MDL_score))
```
